---
title: "Exploratory Data Analysis"
author: "Leona Vracar"
date: "2025-01-12"
output:
  html_document:
    toc: true           # Enables Table of Contents (ToC)
    toc_depth: 3        # Defines ToC depth (1=Main, 2=Subsections, etc.)
    toc_float: true     # Makes ToC float on the side (for easy navigation)
    number_sections: true  # Automatically numbers sections
    code_folding: hide  # Collapses R code chunks by default
    theme: readable     # Sets a readable theme for better visuals
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r EDA Functions}
# Function for analysis of variables
variable_analysis <- function(vect) {
  result <- list(
    mean = mean(vect),
    median = median(vect),
    variance = var(vect),
    sd = sd(vect),
    IQR = IQR(vect),
    min = min(vect),
    max = max(vect),
    range = max(vect) - min(vect)
  )
  return(result)
}

# Just for nicer display in markdown 
# Execute this on a _result variable (output of variable_analysis(swiss$Fertility)) in a code block to get the  nice table in text)
print_nice <- function(result) {
  # put stats in tble
  table <- as.data.frame(t(as.data.frame(result)))
  colnames(table) <- "Value"  # value is 2nd column
  table <- round(table, 4)  
  
  # Print, left align and title
  knitr::kable(table, caption = "Summary Statistics", align = c("l", "r"))
}

```

```{r plots}
# Function for 4 plots: box, histo, QQ, ECDF
plots_function <- function(input_vect, var_name = "default") {
  # mfrow sets 2x2 layout and mar sets plot margins
  par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1)) 
  
  #Boxplot
  boxplot(
    input_vect,
    main = paste("Boxplot of", var_name),
    col = "lightgreen",
    horizontal = TRUE 
  )
  
  # Histogram
  hist(
    input_vect,
    main = paste("Histogram of", var_name),
    xlab = var_name,
    col = "lightblue",
    freq = FALSE
  )
  lines(density(input_vect), col = "red", lwd = 2)  #density curve
  
  # Q-Q 
  qqnorm(
    input_vect,
    main = paste("Q-Q Plot of", var_name),
    col = "blue"
  )
  qqline(input_vect, col = "red", lwd = 2)  
  
  # ECDF
  plot(
    ecdf(input_vect),
    main = paste("ECDF of", var_name),
    xlab = var_name,
    ylab = "ECDF"
  )
  
  # Reset layout back to normal 1x1
  par(mfrow = c(1, 1))
}

```

```{r outliers}
# Function for outliers check using 1.5 × IQR 
calculate_outliers <- function(vect) {
  Q1 <- quantile(vect, 0.25)
  Q3 <- quantile(vect, 0.75)
  IQR <- Q3 - Q1
  result <- list(
    low_fence = Q1 - 1.5 * IQR,
    up_fence = Q3 + 1.5 * IQR
  )
  return(result)
}
  
```

```{r scatter}
# function for scatterplot matrix
scatter_matrix <- function(data, main_title = "Scatterplot Matrix", point_color = "darkblue", point_size = 0.8) {
  pairs(
    data,
    main = main_title,
    col = point_color,
    cex = point_size
  )
}
```

```{r correlation}
# Function for correlation matrix
cor_matrix <- function(data) {
  # Compute cor matrix
  result <- cor(data, use = "pairwise.complete.obs")
  # Print rounded matrix
  print(round(result, 2))
  
  # make heatmap with dendrogram/colors
  heatmap(
    result, 
    main = "Correlation Matrix", 
    col = colorRampPalette(c("blue", "white", "red"))(20)
  )
}
```

# EDA: Dataset "swiss"

The dataset swiss contains 6 variables (columns) and 47 rows
representing 47 French-spaking provinces of Switzerland at 1888.

The dataset provides information about socioeconomic indicators and
fertility for each province.

```{r swiss}
data(swiss)
head(swiss)
```

## EDA of individual variables

In this section, we perform EDA on variables "Fertility", "Agriculture",
"Education", "Catholic" and "Infant Mortality".

```{r EDA swiss all variables}

fertility_result <- variable_analysis(swiss$Fertility)
agriculture_result <- variable_analysis(swiss$Agriculture)
education_result <- variable_analysis(swiss$Education)
catholic_result <- variable_analysis(swiss$Catholic)
infant_result <- variable_analysis(swiss$Infant.Mortality)
```

### Fertility

```{r}
print_nice(fertility_result)
```

**Location**: The variable Fertility has a mean of
**`r round(fertility_result$mean, 2)`** and a median of
**`r round(fertility_result$median, 2)`** suggesting that the data may
be symmetric and that there are likely no (or just a few) outliers that
skew the dateset (otherwise the mean as a non-robust measure would not
so closely match the median).

**Spread/variation**: The data has a moderate spread with a SD of
**`r round(fertility_result$sd, 2)`** which is small relative to the
mean of **`r round(fertility_result$mean, 2)`** and overall range. The
IQR is **`r round(fertility_result$IQR, 2)`** which is small relative to
the range is **`r round(fertility_result$sd, 2)`** suggesting that the
middle 50% of the data points are in a narrow range around the median
(also mean due to mean and median being almost identical).

```{r}
# Call the function
plots_function(swiss$Fertility, var_name = "Fertility")
```

**Boxplot**: In the boxplot, we see **two** potential outliers that lie
beyond the central \~95% of the data. However, these two points may not
necessarily be outliers since the data is left-skewed (longer whiskers
on the left and the location of the median in the left half of the box
suggest this skew).

**Histogram**: From the histogram we clearly see the data is not
symetrical, is **left-skewed** and unimodal with a single peak. This
means that there are potential outliers in the left tail. Most of the
data is concentrated around one SD above and below the mean (two largest
bars are at 60-80).

**Q-Q Plot**: The left-skew caused by potential outliers is also visible
in the Q-Q plot as there are data points deviating from the qqline
(normal distribution) at the left end of the plot.

**ECDF**: The steep slope of the ECDF plot around 0.5 (which aligns with
the median) also shows that most data-points are concentrated around
this value. The curve has a left tail, again due to the two values
outside the 95% of the data, one of which is the minimum value of the
range at **`r round(fertility_result$min, 2)`**.

In the context of the dataset, it seems that fertility was mostly
moderate to high with a few extremes in the lower end of the
distribution that could be worth looking into, especially in correlation
with other variables.

### Agriculture

```{r}
print_nice(agriculture_result)
```

**Location**: The mean at **`r round(agriculture_result$mean, 2)`** and
median at **`r round(agriculture_result$median, 2)`** are roughly close
together suggesting that the data is close to symmetrical.

**Spread/Variation**: The data has a wider spread compared to fertility
with a SD of **`r round(agriculture_result$sd, 2)`** and a wide range of
**`r round(agriculture_result$range, 2)`** from the min value at
**`r round(agriculture_result$min, 2)`** to max value at
**`r round(agriculture_result$max, 2)`**. The middle 50% of data or the
IQR is at **`r round(agriculture_result$IQR, 2)`**, suggesting that the
middle 50% of data is relatively concentrated but with more spread out
values beyond the IQR indicated by the SD of
**`r round(agriculture_result$sd, 2)`**.

**Skewness**: While the mean and median are pretty close together, the
mean is smaller than the median meaning that there might be a left skew
in the data pulling the mean slightly down from the central value.

```{r}
plots_function(swiss$Agriculture, var_name = "Agriculture")
```

**Boxplot**: The length of the left whisker is longer, suggesting a
slight left skew. There are no visible outliers in the box plot.

**Histogram**: The slight left skew can also be seen in the histogram.
The dataset is unimodal and more widely spread out compared to
fertility.

**Q-Q Plot**: The data seems to be very close to a normal distribution,
with most points on the qq-line and a few deviations on the left and
right end.

**ECDF**: The ECDF-curve has a moderately steep slope, suggesting a
wider range.

In the context of the dataset, the lowest value for agriculture
**`r round(agriculture_result$min, 2)`** corresponds to the province "V.
De Geneve" which is to be expected as this was an urban center with
likely little to no agricultural activities.

#### Extra: Shapiro-Wilks test to check normality

For the null hypothesis, we assume data is approximately normally
distributed, while the alternative hypothesis is that data is not
normally distributed.

```{r}
shapiro.test(swiss$Agriculture)
```

With a p-value of 0.193, we fail to reject the null hypothesis.
Agriculture is from a normal distribution.

### Education

```{r}
print_nice(education_result)
```

**Location**: The mean at **`r round(education_result$mean, 2)`** is
above the median at **`r round(education_result$median, 2)`** suggesting
data points which are pulling the mean up from the central value.

**Spread/Variation**: The IQR tells us that 50% of data is in the range
of IQR = **`r round(education_result$IQR, 2)`**. Such a tight IQR
relative to the min (**`r round(education_result$min, 2)`**) and max
(**`r round(education_result$max, 2)`**) values suggest that half the
data is very tightly clustered. The SD of
**`r round(education_result$sd, 2)`** is larger than the IQR meaning the
SD is inflated by a few larger values.

**Skewness**: As stated, both the mean and SD (as non-robust measures)
are inflated upwards compared to the median and IQR (their robust
counterparts) which suggests a right skew or a few large outliers.

```{r}
plots_function(swiss$Education, var_name = "Education")
```

**Boxplot**: The boxplot reflects the tight clustering of the data in
the IQR (the box) as well as a right skew due to the longer right
whisker and four outliers.

**Histogram**: From the density curve, the data looks bimodal with the
first peak at the mean and a second one around 30.

**Q-Q Plot**: The higher values deviating from the qqline are also
visible on the Q-Q plot. While we only identified 4 in the boxplot,
there seem to be several more (6-7) visible in the Q-Q plot. The data
does not follow a normal distribution.

**ECDF**: The ECDF is interesting as it shows a steep curve between \~1
and \~15 on the x-axis, with \~15 corresponding to 0.8 on the y-axis,
meaning that \~80% of data is in this range (low variability). The long
tails at \~20, \~30 and \~50 on the x-axis are extreme values causing
the right-skew. In the context of the dataset, this means that most
people (in 1888) had a low-level of education with just a few having
higher levels. The max value of **`r round(education_result$max, 2)`**
matches the min value for agriculture and corresponds to the province
"V. De Geneve" which likely had a more educated population due to being
an urban center.

### Catholic

```{r}
print_nice(catholic_result)
```

**Location**: The mean at **`r round(catholic_result$mean, 2)`** and
median at **`r round(catholic_result$median, 2)`** differ significantly
which means the data is not symmetrical. Since the mean \>\> median,
there are likely several extreme data points in the higher end of the
distribution affecting the mean.

**Spread/Variation**: The IQR as a robust measure at
**`r round(catholic_result$IQR, 2)`** and the high SD as a non-robust
one at **`r round(catholic_result$sd, 2)`** signify a broad distribution
and highly variable data. This is further confirmed by broad range
(**`r round(catholic_result$range, 2)`**) between min at
**`r round(catholic_result$min, 2)`** and max at
**`r round(catholic_result$max, 2)`**.

**Skewness**: The data is likely bimodal or right-skewed due to the mean
being significantly higher than the median.

```{r}
plots_function(swiss$Catholic, var_name = "Catholic")
```

**Boxplot**: The median is very close to the lower quartile (Q1) in the
box meaning that most data points are likely in the lower range. There
are no visual outliers but the distribution is very stretched out and
the right whisker together with the position of the median suggest a
right skew.

**Histogram**: The histogram has 2 peaks i.e. a bimodal distribution.
Very few values are in the central part of the range (between 30 and
70). The density curve may be a bit misleading because it doesn't show
the sharp peaks visible in the bars.

**Q-Q Plot**: The data is clearly bimodal. The tail in the low end
(negative quantiles on the x-axis) is below the normalcy line which
suggests smaller values compared to a normal distribution. Most data
points are in this lower end with the median (at 0 quantile on the
x-axis) being significantly below the expected median value for a normal
distribution. The higher end of the data (between \~0.5-2 quantiles on
x-axis) show the higher end of the bimodal distribution.

**ECDF**: This plot confirms that the data is bimodal with most values
(\~0.6 on y-axis or 60%) being in a narrow lower range, a few dispersed
values in the mid-range and another distinct peak (\~0.75 on y-axis or
at 75%) in the high range.

In the context of the dataset, the bimodal distribution means that most
provinces were either completely non-Christian or close to 100%
Christian. However, we do not have the data about the population size of
each province so we cannot make any statements about the the absolute
number of Christians or non-Christians per province. For example, the
max value of **`r round(catholic_result$max, 2)`** corresponding to the
province Herens makes it seem like 100% of the inhabitants are Christian
but there may only be a very small population living in this province.

### Infant mortality

```{r}
print_nice(infant_result)
```

**Location**: The mean (**`r round(infant_result$mean, 2)`**) and median
(**`r round(infant_result$mean, 2)`**) are almost identical, suggesting
symmetrical data with little to no outliers.

**Spread/Variation**: Both the SD at **`r round(infant_result$sd, 2)`**
and IQR at **`r round(infant_result$IQR, 2)`** suggest low variability
of data with most data points centered around the mean/median. However,
since SD is slightly lower than the IQR and the data ranges from
**`r round(infant_result$min, 2)`** (min) to
**`r round(infant_result$max, 2)`** (max), there may be some low values
present that slightly push the SD down.

**Skewness**: No indication of skewness.

```{r}
plots_function(swiss$Infant.Mortality, var_name = "Infant Mortality")
```

**Boxplot**: The boxplot shows an outlier which is the min value of
**`r round(infant_result$min, 2)`** corresponding to the province La
Vallee.

**Checking the min value as potential outlier**

```{r}
infant_outliers <- calculate_outliers(swiss$Infant.Mortality)
infant_outliers
```

Using the Q1 - IQR \* 1.5 (lower fence at
**`r round(infant_outliers$low_fence,  2)`**) and Q3 + IQR \* 1.5 (upper
fence at **`r round(infant_outliers$up_fence, 2)`**), we see that the
min value lies outside this range and is technically an outlier.
However, this entry is worth examining further as a data point with some
additional significance (also in relation to other variables).

**Histogram**: The data shows a normal distribution except for the
aforementioned outlier/min value.

**Q-Q Plot**: This plot also confirms normalcy of the data as the
individual points are on the qqline.

**ECDF**: Also shows normal distribution of data excluding the min
value.

## Correlation between variables

**Scatterplot Matrix**

```{r}
# Selecting variables (making a subset of those that we're interested in)
swiss_subset <- swiss[, c("Fertility", "Agriculture", "Education", "Catholic", "Infant.Mortality")]

# Scatterplot matrix
scatter_matrix(swiss_subset)
```

The scatterplot matrix shows visible patterns between some variables:

1.  Fertility and Agriculture seem to have a positive correlation.
2.  Fertility and Education seem to have a negative correlation.
3.  Infant mortality and Fertility seem to have a positive correlation.
4.  Agriculture and Education seem to have a negative correlation.

To examine these correlations further, we look at a correlation matrix
in the next step.

**Correlation Matrix**

```{r}
cor_matrix(swiss_subset)
```

The correlation matrix confirms the correlations we already identified
and additionally shows a positive correlation between Fertility and
Catholic (r = 0.46).

Some possible conclusions from the correlations are:

**Fertility, Agriculture and Education**: Swiss provinces with higher
rates of agricultural activity had a higher fertility rate while
education and fertility were negatively correlated (r = -0.66). In
addition, education and agriculture were strongly negatively correlated
(r = -0.64), suggesting that education levels were lower in rural areas
and higher in urban areas. These findings are in line with common
demographic patterns where lower birth rates are associated with higher
education levels.

**Fertility and Infant Mortality**: Higher fertility rates also have a
positive correlation with infant mortality (r = 0.46). One (speculative)
reason for this may be that having more children automatically leads to
more children not surviving into adulthood. We could also speculate,
that since fertility is higher in rural areas i.e. positively correlated
with agriculture, it is possible that in 1888, medical care was not
available which lead to more infant deaths.

**Catholic**: One somewhat peculiar finding was the positive correlation
between Catholic and Fertility (r = 0.46) suggesting that Catholic
provinces had higher fertility rates. Catholicism rates also had a
positive correlation with Agriculture (r = 0.40) suggesting that more
agricultural communities were more religious and had a higher rate of
fertility.

# EDA: Dataset "state.x77"

The state.x77 dataset includes various statistics (population, income,
illiteracy etc.) from the 1970s for each USA-state.

```{r}
state.x77_dataframe <- as.data.frame(state.x77)
head(state.x77_dataframe, 7)
```

## EDA of individual variables

This section explores the variables "Population", "Income",
"Illiteracy", "Life.Exp", "Murder", "HS Grade" and "Frost".

```{r}
pop_result <- variable_analysis(state.x77_dataframe$Population)
inc_result <- variable_analysis(state.x77_dataframe$Income)
illi_result <- variable_analysis(state.x77_dataframe$Illiteracy)
life_exp_result <- variable_analysis(state.x77_dataframe$`Life Exp`)
murder_result <- variable_analysis(state.x77_dataframe$Murder)
hs_grad_result <- variable_analysis(state.x77_dataframe$`HS Grad`)
frost_result <- variable_analysis(state.x77_dataframe$Frost)
```

### Population

```{r}
print_nice(pop_result)
```

**Location**: The mean at **`r round(pop_result$mean,  2)`** and median
at **`r round(pop_result$median,  2)`** diverge significantly which
suggests the data is not symmetrical. Median is robust and hence
appropriate as the locus for this dataset.

**Spread**: The data has a high variance
**`r round(pop_result$variance,  2)`** and standard deviation
**`r round(pop_result$sd,  2)`** which means the population numbers are
widely different across states. The IQR of 3889 means that the middle
50% of states have populations in this range.

**Skewness**: The data is skewed to the right since mean \>\> median.

```{r}
plots_function(state.x77_dataframe$Population, var_name = "Population")
```

**Boxplot**: Outliers are visible beyond the right whisker, 3 potential
outliers and 2 definite ones. The elongated right whisker also suggests
a right skew.

**Histogram**: The right-skew is clearly visible in the histogram. Most
states have a population up to 7000, while there are a few ones with
larger population numbers between 10 000 and 25 000. It could be argued
that the data is unimodal with several outliers or the outliers could be
grouped together in a cluster in a bimodal dataset since population is
(unsurprisingly) higher in more urban states like California and New
York.

**Q-Q Plot**: The dataset deviates from a normal distribution,
especially at the upper end. Again, the deviation is caused by the
outliers.

**ECDF**: The ECDF plot shows a steep rise in the beginning, meaning
many states have low populations. The more elongated points at the right
of the plot are the few states with large populations.

### Income

```{r}
print_nice(inc_result)
```

**Location**: The median at **`r round(inc_result$median,  2)`** and
mean at **`r round(inc_result$mean,  2)`** are close together, hence the
data is likely symmetrical. From the plots below, there seems to be a
single potential outlier in the higher range of incomes but the median
is still a fitting measure of central tendency for this variable.

**Spread**: The per capita income has a moderate spread with a sd of
**`r round(inc_result$sd,  2)`**. The robust measures: range of
**`r round(inc_result$range,  2)`** and IQR of
**`r round(inc_result$IQR,  2)`**.

**Skewness**: The data approximately follows a normal distribution.

```{r}
plots_function(state.x77_dataframe$Income, var_name = "Income")
```

**Boxplot**: A single outlier is visible in the high range. The central
50% of per-capita incomes are between 4000 and 5000 USD.

**Histogram**: The data is unimodal and approximately normally
distributed.

**Q-Q Plot**: The normal distribution is also confirmed by the Q-Q plot.

**ECDF**: Smooth, S-shaped curve, indicating a gradual increase in
values. Most incomes are between \~3500 and \~5000.

### Illiteracy

```{r}
print_nice(illi_result)
```

**Location**: Both the mean at **`r round(illi_result$mean,  2)`** and
the median at **`r round(illi_result$median,  2)`** are low and pretty
close together without any outliers. While the mean is appropriate here
as a measure of central tendency, it is slightly higher than the median,
potentially indicating a slight right-skew.

**Spread**: Considering that the illiteracy rate is a percentage of the
population of the respective state (and hence could technically take up
values 0-100%), the data has low variability and a tight spread. The min
at **`r round(illi_result$mean,  2)`** and max at
**`r round(illi_result$max,  2)`** confirm this. At most, the illiteracy
rate is at 2.8% which is indicative of the fact that the US population
across all states was highly literate in the 1970s.

**Skewness**: Illiteracy is slightly right-skewed, meaning most states
had low illiteracy, but a few had significantly higher rates as visible
in the right whisker of the boxplot and the density curve of the
histogram. However, it is crucial to interpret the data in its context
here. Considering that illiteracy could take up values between 0-100%
and the right-skew is caused by a few values between \~ 1.6% illiteracy
(displayed by the right whisker in the box plot) and the max value of
2.8%, illiteracy still remains very low overall.

```{r}
plots_function(state.x77_dataframe$Illiteracy, var_name = "Illiteracy")
```

**Boxplot**: There are no extreme outliers, suggesting no states with
high illiteracy rates.

**Histogram**: The highest bar contains values between 0.5 and 1%. The
data has a right skew and is unimodal.

**Q-Q Plot**: The data is approximately normally distributed with a
slight deviation in the bottom range.

**ECDF**: Around 60% of states have illiteracy below 1.5%, as seen by
the more steep rise before 1.5.

### Life Expectancy

```{r}
print_nice(life_exp_result)
```

**Location**: The mean at **`r round(life_exp_result$mean,  2)`** and
median at **`r round(life_exp_result$median,  2)`** are almost identical
which suggests highly symmetrical data.

**Spread**: The standard deviation at
**`r round(life_exp_result$sd,  2)`** is \~23% of the total range
(**`r round(life_exp_result$range,  2)`**) suggesting a tight to
moderate spread relative to the range. The IQR of
**`r round(life_exp_result$IQR,  2)`** as a robust measure confirms
this. In the context of life expectancy, most values are concentrated
around the mean.

**Skewness**: Data appears to be symmetrical with no outliers nor
skewness.

```{r}
plots_function(state.x77_dataframe$`Life Exp`, var_name = "Life Exp")
```

**Boxplot**: No outliers and approximately symmetrical data.

**Histogram**: Data appears to be unimodal and approximately normally
distributed. Three tallest bars are in the 70-73 range, which sum up to
\~75% on the density on the y-axis. Hence, a large majority of states
have life expectancies in this range.

**Q-Q Plot**: This plot confirms that life expectancy follows a normal
distribution with only a few minor deviations at the upper/lower end.

**ECDF**: Pretty smooth curve which gets steeper at LifeExp = 70 and
then flattens out again around LifeExp = 73. This confirms that the top
\~75% of data is located in this range.

### Murder Rate

```{r}
print_nice(murder_result)
```

**Location**: The mean (**`r round(murder_result$mean,  2)`**) is
slightly higher than the median
(**`r round(murder_result$median,  2)`**), indicating mild right
skewness. The mean may still be appropriate as the central tendency of
the data.

**Spread**: The range (**`r round(murder_result$range,  2)`**) is pretty
large relative to the central tendency, suggesting a wider spread of
data.

**Skewness**: As visible in the plots, the data is not skewed but
bimodal with a second peak which is displacing the mean upwards compared
to the median.

```{r}
plots_function(state.x77_dataframe$Murder, var_name = "Murder")
```

**Boxplot**: No visible outliers and approximately symmetrical data. The
right whisker is longer which means that more values are found in the
upper 25% of data compared to the lowest 25%.

**Histogram**: There are two clear peaks in the histogram, one between
Murder \~3 to \~6.5 and a second one between \~10 and \~12. Hence,
murder appears to be bimodal with \~30% of states exhibiting a higher
murder rate per 100 000 inhabitants compared to the rest of the data
contained in the first mode which have murder rates lower than 10.

**Q-Q Plot** and **ECDF**: The bimodality of the data is visible in both
the Q-Q plot and the ECDF. The ECDF is particularly interesting as it
clearly shows a gradual rise of the curve from around 0 to 7, then a
brief plateau follower by a second steep rise in the 10-12 range. The
Q-Q plot shows significant deviations from normality beyond 10 on the
y-axis (sample quantlies) which corresponds to the second peak in the
bimodal distribution.

In the context of the dataset, murder rates seem to be surprisingly high
in some states compared to the rest of the country.

### HS Graduation Rate

```{r}
print_nice(hs_grad_result)
```

**Location**: The mean (**`r round(hs_grad_result$mean,  2)`**) and
median **`r round(hs_grad_result$median,  2)`** are approximately
identical, which in isolation suggests the data may be highly
symmetrical. However, the standard deviation at
**`r round(hs_grad_result$sd,  2)`** is high compared to the IQR of
**`r round(hs_grad_result$IQR,  2)`** which indicates an unusually high
variability in the data and potential bimodality.

**Spread**: The data has a wide spread between min =
**`r round(hs_grad_result$min,  2)`** and max =
**`r round(hs_grad_result$max,  2)`** (range =
**`r round(hs_grad_result$range,  2)`**).

**Skewness**: No indication of skewness.

```{r}
plots_function(state.x77_dataframe$`HS Grad`, var_name = "HS Grad")
```

**Boxplot**: The boxplot, on its own, suggests almost perfect symmetry
but the other plots clearly indicate bimodality.

**Histogram**: The histogram reveals two peaks: a smaller one in the
lower range and a larger peak in the higher range. This means that the
percentage of high school graduates in most states is between \~50-60%.
However, a smaller but non-negligible number of states have lower high
school graduation rates in the \~38-43% range.

**Q-Q Plot** and **ECDF**: The Q-Q plot shows that the data strongly
deviates from a normal distribution, especially in the lower range which
corresponds to the first mode. The ECDF illustrates the separation of
the two peaks: a short but steep curve of the first mode is followed by
a plateau between \~43 and \~46 on the x-axis, followed by a second
curve with a more gradual increase but a small steep rise at \~53. The
small steep rise at \~53 on the ECDF curve signifies that there is a
concentration of data at this point which may be part of the reason why
the mean also has this value and wasn't displaced from the median by the
bimodal distribution.

### Frost

The frost variable is the mean number of days with minimum temperature
below freezing (1931-1960) in capital or large city.

```{r}
print_nice(frost_result)
```

**Location**: The mean (**`r round(frost_result$mean,  2)`**) and
(**`r round(frost_result$median,  2)`**) differ by 10 days which
suggests that there are low values displacing the mean below the median.
The mean approximates the central tendency of the data but the median is
more reliable as the locus in this case.

**Spread**: The standard deviation (**`r round(frost_result$sd,  2)`**)
is large relative to the overall range of the data
(**`r round(frost_result$range,  2)`**), meaning that the data is highly
spread out and variable.

**Skewness**: From the histogram and Q-Q plot we can conclude that the
data is actually bimodal with some states having low numbers of days
with freezing temperatures (e.g. Hawaii, our min value of 0).

```{r}
plots_function(state.x77_dataframe$Frost, var_name = "Frost")
```

**Boxplot**: The boxplot is not a useful visualization for this data as
it does not show the bimodal distribution. The only useful information
we can deduce is that the slightly longer left whisker means there are
several states with little to no days with below-freezing temperatures.

**Histogram**: The density curve is misleading as it smooths out the
first peak of the bimodal distribution i.e. the data contained in the
first bar. However the histogram bars are the best visualization here
for the bimodallity of the data.

**Q-Q Plot**: The Q-Q plot on its own it would not clearly indicate
bimodality as most points fall close to the red qqline. However, the
typical S-shaped curve of a bimodal distribution is still visible in the
Q-Q plot, albeit its S-shape is more flattened out and may be difficult
to notice at first glance.

**ECDF**: Similarly, the ECDF plot barely makes the bimodal distribution
visible. It is possible to identify a clear second mode around 130 on
the x-axis but the first mode is barely distinguishable between 5 and
10.

## Correlation between variables

### Scatterplot Matrix

```{r}
usa_subset <- state.x77[, c("Population", "Income", "Illiteracy", "Life Exp", 
                            "Murder", "HS Grad", "Frost")]
# Scatterplot matrix
scatter_matrix(usa_subset)
```


### Correlation Matrix


```{r}
cor_matrix(usa_subset)
```


In this part, we look at some interesting correlations between the
variables:

**Illiteracy and Murder Rate (0.70)**: A strong positive correlation
suggests that states with higher illiteracy rates tend to have higher
murder rates. This indicates that education levels likely have something
to do with murder rates or crime in general.

**Illiteracy and Life Expectancy (-0.59)**: The negative correlation
implies that states with higher illiteracy rates tend to have lower life
expectancy. We can speculate that illiteracy likely doesn't exist in
isolation but that it is the result of generally worse economic
conditions and less access to things such as healthcare, balanced food
or even free time to learn to read and write. Hence, illiteracy probably
doesn't lower life expectancy on its own but it signifies a larger issue
like poverty in these states which leads to lower life expectancy and
more crime.

**Murder Rate and Life Expectancy (-0.78)**: A strong negative
correlation indicates that states with higher murder rates have lower
life expectancy, which is not surprising and maybe shows that there are
public safety issues in these states.

**HS Graduation Rate and Life Expectancy (0.58)**: A positive
correlation suggests that states with higher high school graduation
rates also tend to have longer life expectancy. Again, we cannot make
the conclusion that graduating high school increases life expectancy but
the positive correlation may point to other societal issues which cause
people to not graduate high school such as worse living conditions which
in turn may lower life expectancy.

**Income and HS Graduation Rate (0.62)**: A moderately strong positive
correlation suggests that states with higher average income levels tend
to have higher high school graduation rates, which makes sense as richer
states may have more funding for education, more people can afford
education and access it more easily.

**Frost and Murder Rate (-0.54)**: A negative correlation suggests that
colder states (those with more frost days) tend to have lower murder
rates.

**Population and Murder Rate (0.34)**: A weak positive correlation
suggests that more populous states tend to have slightly higher murder
rates maybe because crime is often concentrated in big cities.

# EDA: Dataset "LakeHuron"

The dataset LakeHuron contains annual measurements of the level of Lake
Huron 1875-1972 (in feet).

```{r}
LakeHuron <- LakeHuron
```

## EDA of lake levels

### Lake levels without considering the time aspect

```{r}
# To analyze just the lake levels we convert them to a numeric vector
# LakeHuron is a ts object in R
lake_levels <- as.numeric(LakeHuron)

levels_result <- variable_analysis(lake_levels)
print_nice(levels_result)
```

**Location**: The mean (**`r round(levels_result$mean,  2)`**) and
median (**`r round(levels_result$median,  2)`**) are close to being
identical which suggests a high level of symmetry and no outliers or
extreme values.

**Spread**: The data has a moderate range
(**`r round(levels_result$range,  2)`**) and moderate spread. The middle
50% of data is also pretty concentrated with an IQR of
**`r round(levels_result$IQR,  2)`**.

**Skewness**: The mean is slightly lower than the median which suggests
that a few values below the median may be causing a slight deviation
from symmetry. However, the data still appears to be very close to
perfect symmetry.

```{r}
plots_function(lake_levels, var_name = "Lake Levels (no time)")
```

**Boxplot**: A few more data points below the median (the black line)
can be deduced from the boxplot. Besides this, the data seems symmetric
with no outliers.

**Histogram**: Data is approximately symmetric and unimodal with a very
slight left tail but not nearly enough to identify skewness.

**Q-Q Plot**: Data is normally distributed with just a few deviations
from the red qqline at the extremes.

**ECDF**: The ECDF is smooth and continuous, indicating no outliers.

### Lake levels including the time aspect

From the time series plot above, there seems to be a decline in water
levels as time progresses. However, we can't say if this decline is
significant at this stage of analysis.

Another interesting element in this data is that the max value was
recorded in the first few years of measurements while the lowest water
level was recorded near the end around 1965.

```{r}
plot(LakeHuron, main = "Lake Huron Water Levels (1875–1972)",
     xlab = "Year", ylab = "Water Level", col = "blue", type = "l", xaxt = "n")
axis(1, at = seq(1875, 1975, by = 5), labels = seq(1875, 1975, by = 5))

# Min and max values
points(time(LakeHuron)[which.min(LakeHuron)], levels_result$min, col = "red", pch = 19)
points(time(LakeHuron)[which.max(LakeHuron)], levels_result$max, col = "red", pch = 19)

```

Upon examining the lowest recorded values (the valleys on the plot), it
appears that over time the lowest recorded water level was being pushed
further and further down.

# Titanic

Task: **Explore and visualize the Titanic dataset. How do age, gender,
and class influence survival? Identify where Simpson's Paradox appears
and explain the reasons behind this effect.**

## Data Exploration

```{r}
str(Titanic)
summary(Titanic)
mosaicplot(Titanic, main ="Survival on the Titanic", color = TRUE)
apply(Titanic, c(1,2,3,4), sum)
```

The dataset "Titanic" represents the survivors/non-survivors of the
Titanic. They are divided into Sex (female-male), Class (1st, 2nd, 3rd,
Crew), Age (Child, Adults) and Survival (Yes, No). An overview of this
data can be seen in the mosaic plot.

## Impact of Age, Sex and Class on Survival

```{r, echo=FALSE}

# Titanic-Datensatz als Tabelle verwenden
titanic <- as.table(Titanic)

layout(matrix(1:4, ncol = 2))
# 1. Einfluss von Alter auf Überleben
apply(titanic, c(3, 4), sum)
mosaicplot(~ Age + Survived, titanic, color = TRUE, main = "Age vs Survived")

# 2. Einfluss von Geschlecht auf Überleben
apply(titanic, c(2, 4), sum)
mosaicplot(~ Sex + Survived, titanic, color = TRUE, main = "Sex vs Survived")

# 3. Einfluss von Klasse auf Überleben
apply(titanic, c(1, 4), sum)
mosaicplot(~ Class + Survived, titanic, color = TRUE, main = "Class vs Survived")

# 4. Einfluss von Klasse, Geschlecht und Überleben
apply(titanic, c(1, 2, 4), sum)
mosaicplot(~ Class + Survived + Sex, titanic, color = TRUE, main = "Class, Sex vs Survived")

layout(1)

# 4. Einfluss von Alter, Klasse und Überleben
apply(titanic, c(1, 2, 4), sum)
mosaicplot(~ Class + Survived + Age, titanic, color = TRUE, main = "Age, Class vs Survived")


```

## Simpsons paradox & Interpretation

Simpsons Paradox is a statistical phenomenon that occurs when subgroups
are combined into one group. The process of aggregating data can cause
the apparent direction and strength of the relationship between two
variables to change.

In the case of our Titanic dataset one would think that the survival
rate of the female crew was very low. But if we look at the "Sex vs.
Survived" plot in general the females had a better survival outcome.
This is because there were not many female crew members, but overall 20
out of 23 female crew members survived.

Furthermore if we only look at "Class vs. Survived" one could think,
that the survival is independent from sex - so the the better the class,
the better the survival outcome. But if we also include the sex, we see
for example that the survival rate of females in the first class is
almost 100%, whereas men in the first class only had a survival rate of
about 34%.

Then if we look closely at the "Children in the Age vs. Survived" plot,
we may think children do have a higher survival rate than adults. If we
also include the class it becomes clear, that the survival rate of
children is also very dependent on the class. The higher the class, the
higher the survival rates in children.
