---
title: 'Regression Modelling'
author: "Leona Vracar"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: readable
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Function for analysis of variables
variable_analysis <- function(vect) {
  result <- list(
    mean = mean(vect),
    median = median(vect),
    variance = var(vect),
    sd = sd(vect),
    IQR = IQR(vect),
    min = min(vect),
    max = max(vect),
    range = max(vect) - min(vect)
  )
  return(result)
}

print_nice <- function(result) {
  # put stats in tble
  table <- as.data.frame(t(as.data.frame(result)))
  colnames(table) <- "Value"  # value is 2nd column
  table <- round(table, 4)  
  
  # Print, left align and title
  knitr::kable(table, caption = "Summary Statistics", align = c("l", "r"))
}

```

# 1: United Nations statistics

In general the dataset UN provides information about National Statistics
from the United Nations, Mostly from 2009–2011.

The dataset contains 7 variables (columns) which are region, group,
fertility, per capita gross domestic product, female life expectancy,
percent urban and infant mortality and 213 rows representing places
(mostly UN members).

```{r}
library("car")
data("UN")
```

## Filtering of 'NA' values

To filter 'NA' values the function `na.omit` is used. This leaves us
with:

```{r}
#'NA' Werte entfernen mit na.omit
cleaned_UN <- na.omit(UN)
summary(cleaned_UN)
```

## Plotting the given data

Infant mortality is to be explained by gross domestic product. For these
data, a scatterplot and residual plots are first created to check all
the necessary statistical assumptions for the validity of the linear
model.

```{r plots }
#Scatterplot for Infant Mortality and gross domestic product
plot(cleaned_UN$ppgdp, cleaned_UN$infantMortality)

#Residual plots for Infant Mortality and gross domestic product
summary(lm(formula = infantMortality ~ ppgdp, data = cleaned_UN))
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm(formula = cleaned_UN$infantMortality ~ cleaned_UN$ppgdp, data = UN))


```

## Log transformation of the data & model fitting

The data cannot be directly fitted with a linear model. The adjusted
R-squared for example is only 0.2618. This is a power function (there
are heavy tails on both sides). Therefore, a log transformation is
applied to both axes.

```{r}
#log10 transformation of y-values + x-values
log_infant_Mortality <- log(cleaned_UN$infantMortality)
log_ppgdp <- log(cleaned_UN$ppgdp)
```

```{r }
#model adaption
model <- lm(log_infant_Mortality~log_ppgdp, data=cleaned_UN)
summary(model)
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(model)
plot(log_ppgdp, log_infant_Mortality)
abline(model)

```

After the transformation we see a linear relationship between the
"infant Mortality" and the "ppgdp" (per capita gross domestic product).
The adjusted R-squared is now at 0,765, which is way better than before
and means that the model explains the data quite well. Furthermore,
looking at the residual plots, we can accept the following assumptions:

-\> The model has no systematic error.

-\> The error variance is the same for all observations (homoscedastic).

-\> The components of the error term are uncorrelated.

-\> The model error is normally distributed.

Therefore we can now fit a linear model.

### Linear Model Equation

$log(InfantMortality) = 8.10377  - 0.61680 * log(gdp)$

We should not forget, that the original data is a power function. So we
have to adapt the linear model equation to be able to interpret the
given data.

$InfantMortality=e^{8.10377} * gdp^{−0.61680}$

This means if the gdp is 1, the infant mortality will be e\^8,10377 (so
about 3306,9 Infant deaths by age one per 1000 live births) and if the
gdp increases by 1%, the infant mortality decreases by 0.6168%.

# 2: Swiss demographic data

The swiss dataset is a data frame with 47 observations on 6 variables
about Fertility and Socioeconomic Indicators from 1888. The 6 variables
are Fertility, Agriculture, Examination, Education, Catholic and Infant
Mortality.

## Mutiple Linear Regression

```{r}
# load the dataset 
data("swiss")
summary(swiss)

# put (absolute) correlations on the upper panels,
# with size proportional to the correlations.
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r, col=ifelse(r>0.5,2,1
  ))
}
#put histograms on the diagonal 
panel.hist <- function(x, ...)
{
  usr <- par("usr")
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}


pairs(swiss, lower.panel = panel.smooth,diag.panel = panel.hist, upper.panel = panel.cor)
summary(lm(formula = Education ~ ., data = swiss))
```

Examination was kicked out of the model, due its high correlation to
Fertility, Agriculture, Catholic and Infant Mortality. It is better to
eliminate one variable than the other four.

## Model adaptation

### Checking the significance of estimates

```{r}
# Filtering Examination 
pairs(swiss[,-3], lower.panel = panel.smooth,diag.panel = panel.hist, upper.panel = panel.cor)
# Model adaption 
lm_model_1 <- lm(formula = Education ~ . - Examination, data = swiss) 
summary(lm_model_1)
```

After fitting a linear regression model predicting Education based on
all variables in the swiss dataset (Examination excluded), we see that
the p value (**0.28**) of infant mortality is not significant. Hence,
infant mortality is removed from the model.

```{r}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_1)
par(mfrow = c(1, 1))
# kicked out: Infant Mortality 
summary(lm(formula = Education ~ . - Examination - Infant.Mortality, data = swiss))
```

With infant mortality and Examination removed, the intercept changes to
**53.8505**, showing that this value has increased compared to
**49.99303** intercept with the infant mortality in the dataset. It is
also apparent, that the estimates for fertility and agriculture
decreased to **-0.48** and **-0.23**, whereas the estimate for the
Catholic variable increased above **0.0833** when infant mortality is
kicked out of the dataset.

Overall, the Multiple R-squared with Infant Mortality in the dataset has
a value of \*\*0.7305\*\* and without \*\*0.7048**. The Adjusted
R-squared changes from** 0.7048 **to** 0.7055\*\* without this variable.
Thus, it can be observed that infant mortality can be excluded from the
model as it doesn't significantly decrease the R-squared value.

Therefore, we now have a multiple model that includes Fertility,
Agriculture, and Catholic. With these, the residual plots are now
created:

```{r}
# Model adaption without Infant Mortality 
lm_model_2 <- lm(formula = Education ~ . - Examination - Infant.Mortality, data = swiss)

par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_2)
par(mfrow = c(1, 1))
```

-\> The model has no systematic error.

-\> The error variance is the same for all observations (homoscedastic).

-\> The components of the error term are uncorrelated.

-\> The model error is normally distributed.

Therefore the fit of a linear model is correct.

### Leverage Point

As seen in plot 1, Val de Geneve has a higher education rate, which can
also be detected in different plots, especially in the leverage plot.
Meaning that this value is a leverage point and we have to look at it
more details. This point is the 45th observation and is kicked out in
the code below, to see if it is a good or bad leverage point.

```{r}
# kicked out Val de Geneve point. 
summary(lm(formula = Education ~ . - Examination - Infant.Mortality, data = swiss[-45, ]))
```

Without Val de Geneve the calculated Intercept was **45.27** compared to
**53.85**. Therefore, we refer to a 10% points poorer fit. The fertility
gradient is now lower than it was within the Val de Geneve in the data
set. In general, the intercept has a great influence on the leverage
point and one can decide between a good or bad leverage point.
Especially the Multiple (0.6216) and Adjusted R-squared (0.5945) show
that the Val de Geneve value is necessary in the data set. As a result,
we speak of a good leverage point, due to its ability to stabilize the
regression line. Despite the fact that it lies far from the other
points, it is important to leave it in the data.

### Summary

In this example the Education (the % education beyond primary school for
draftees) of Swiss in 1888 is explained in a multiple linear regression
model through the variables Fertility, Agriculture and Catholic.

The equation is stated as followed: $$ 
Education = 45.2863 - 0.38468 * Fertility - 0.20240 * Agriculture + 0.06188 * Catholic
$$ This means that if all the other variables are 0, Education beyond
primary school is about 45% for draftees. If Fertility increases by 1
unit, Education decreases by 0.3847 percentage points. If Agriculture
increases by 1 unit, Education decreases by 0.020240 percentage points
and if Catholic is increased by 1 unit, Education is increased by
0,06188 percentage points. So Agriculture and Fertility have a
decreasing effect on Education, but Catholic has an increasing effect on
Education.

# 3: US state demographic data

The USA dataset (state.x77) contains a matrix with 50 rows an 8 columns.
The data is related to the 50 states of the United States of America and
contains information about Population, Income, Illiteracy, Life
Expectancy, Murder, High School graduates as well as Frost and Area.

```{r}
state.x77_dataframe <- as.data.frame(state.x77)
head(state.x77,10)

```

## Multiple Linear Regression

```{r}
# put (absolute) correlations on the upper panels,
# with size proportional to the correlations.
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r, col=ifelse(r>0.5,2,1
  ))
}
#put histograms on the diagonal 
panel.hist <- function(x, ...)
{
  usr <- par("usr")
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}

pairs(state.x77_dataframe, lower.panel = panel.smooth,diag.panel = panel.hist, upper.panel = panel.cor)

```

### Filtering based on Multicolinearity and Estimate Significance

In this step, we look at multicolinearity of the variables and decide
which ones need to be removed in order to correctly fit a multiple
linear regression model. We also check the statistical significance of
the estimates for each variable and remove them if they are below the
0.05 significance level.

```{r}
# Model adaptation
lm_model_1 <- lm(formula = Murder ~., data = state.x77_dataframe) 
summary(lm_model_1)
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_1)
par(mfrow = c(1, 1))

```

Illiteracy and HS Grade were kicked out of the model, due to their
correlation with Income. Kicking out HS grade results in eliminating its
correlation with life expectancy.

```{r}
# Model adaption without Illiteracy and Life Exp
lm_model_2 <- lm(formula = Murder ~. - Illiteracy -`HS Grad`, data = state.x77_dataframe) 
summary(lm_model_2)
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_2)
par(mfrow = c(1, 1))
pairs(state.x77_dataframe[, -c(3, 6)], lower.panel = panel.smooth,diag.panel = panel.hist, upper.panel = panel.cor)

```

The data shows that Income has a p-value of **0.530396** and it not statisticalyl significant.
Therefore, it is removed from the model.

### MLR After Filtering

```{r}
# Model adaption without Illiteracy and Life Exp
lm_model_3 <- lm(formula = Murder ~. - Illiteracy -`HS Grad` -Income , data = state.x77_dataframe) 
summary(lm_model_3)
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_3)
par(mfrow = c(1, 1))

```

All values (Population, Life Exp, Frost and Area) demonstrate
statistical significance.

In the plot **"Residuals vs Leverage"** the point Alaska seems to have
an influence on the graph. Therefore, a model can be calculated without
the value, to identify if its happens to be a leverage point. If so, we
analyse whether its a good or bad leverage point.

### Leverage Point

```{r}
# kicked out: Alaska point 
lm_model_4 <- lm(formula = Murder ~. - Illiteracy -`HS Grad` -Income, data = state.x77_dataframe[-2, ])
summary(lm_model_4)
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
plot(lm_model_4)
par(mfrow = c(1, 1))
```

The plot "Residual vs Leverage" without the Alaska value is different
compared to the plot that includes Alaska. Therefore, the Alaska value
seems to be a bad leverage point as it displaces the red line in the
leverage plot further away from being horizontal. The Multiple and
Adjusted R-Squared did not change a lot. Therefore, this leverage point
is not necessary for this model and can be kicked out.

If we now take a look at the final residual plots, we can say that:

-\> The model has no systematic error.

-\> The error variance is the same for all observations (homoscedastic).

-\> The components of the error term are uncorrelated.

-\> The model error is normally distributed (even if there are to small
"humps" nevertheless we would say that the model error is normally
distributed.

Therefore the fit of a linear model is correct.

## Summary

In this example the Murder rate in the US is explained in a multiple
linear regression model through the variables Population, Life
Expectancy, Frost and Area.

The equation is stated as followed:

$$ 
Murder = 141.4 + 0.000143 * Population - 1.879 * Life Expectation - 0.02135 * Frost + 0.00001246 * Area 
$$ This means that if all the other variables are 0, Murder rate is
about **141.4**. Mathematically, if Population increases by 1 unit,
Murder rate increases by **0.000143** percentage points.

If Life Expectancy increases by 1 unit, Murder rate decreases by
**1.879** percentage points. If Frost is increased by 1 unit, Murder
rate is decreased by **0.02135** percentage points and if Area is
increased by 1 unit, Murder rate is increased by **0.00001246**
percentage points. So Life Population and Area have a very small
positive effect on Murder rate, but Life Expectancy and Frost have a
negative effect on Murder rate. With a positive effect on the murder
rate, we speak of an increase of this rate, whereas a negative effect
stands for a decrease.

# 4: Lake Huron timeseries dataset

LakeHuron is a time series dataset with annual measurements of the level
of Lake Huron 1875-1972 (in feet).

```{r }

lake_levels <- as.numeric(LakeHuron)

levels_result <- variable_analysis(lake_levels)

print_nice(levels_result)

plot(LakeHuron, main = "Lake Huron Water Levels (1875–1972)",
     xlab = "Year", ylab = "Water Level", col = "blue", type = "l", xaxt = "n")

axis(1, at = seq(1875, 1975, by = 5), labels = seq(1875, 1975, by = 5))
```

## Model selection

From the time series plot above, there seems to be a decline in water
levels as time progresses. However, we can't say if this decline is
significant at this stage of analysis. There is no symmetrical periodic
pattern in the data which is expected as the measurements were taken
annually.

Next, we check the correlation between time (years) vs. lake levels and
do a correlation test to check its significance.

```{r}

time <- time(LakeHuron) # Extract the time (years)
cor_time_levels <- cor(time, lake_levels)
cor_time_levels
```

The correlation between time and lake levels is **`r cor_time_levels`**
which suggests that as one variable increases, the other decreases i.e.
as time increases, the lake levels decrease. This correlation
coefficient is in line with the visual observation of lake levels
decreasing over time. It is important to note that since this is
time-series data, there is no actual real-world correlation between the
two variables i.e. time doesn't affect the change in lake levels but
rather the lake levels decreased over time.

To check if this correlation coefficient is significant, we perform both
a two-tailed and one-tailed correlation test. In the two-tailed test,
the null hypothesis is that the correlation coefficient is 0 and the
alternative hypothesis is that the correlation coefficient is not equal
to 0. In the one-tailed test, the null hypothesis is that the
correlation coefficient is greater than or equal to 0 while the
alternative hypothesis is that it is less than 0. Both tests are done at
a 95% confidence level.

```{r}
#two-tailed correlation test
cor_test_two_tailed <- cor.test(time, lake_levels, alternative = "two.sided")
print(cor_test_two_tailed)

#one-tailed correlation test (less)
cor_test_one_tailed <- cor.test(time, lake_levels, alternative = "less")
print(cor_test_one_tailed)
```

The two-tailed test confirms that the correlation value is statistically
significant with a p value of **3.545e-08**. The one-tailed test
produces an even lower p-value of **1.773e-08** which tells us the
negative correlation coefficient is indeed statistically significant.

## Linear Regression Model

Since there are only have 2 variables (time and lake levels) and lake
levels seem to decline in a linear fashion, we fit a simple linear model
to the data.

```{r}
model_lake_huron <- lm(LakeHuron ~ time(LakeHuron))
summary(model_lake_huron)

plot(LakeHuron, main = "Lake Huron Water Levels (1875–1972)",
     xlab = "Year", ylab = "Water Level", col = "blue", type = "l")
abline(model_lake_huron, col = "red", lwd = 1)
```

From the linear model we can extract the following information:

-   **Residuals**: The residuals are the differences between observed
    levels and predicted levels in the regression model. They are are
    symmetric around the median which is very close to 0 and have a
    spread of about ±2.5.

-   **Slope**: The slope is **-0.0242** which means that lake level
    decreases by 0.0242 percentage points feet per year on average with a standard error
    of 0.004036.

-   **Multiple R squared**: 27.25% of the variability in lake levels is
    explained with the linear trend over time. This score is pretty bad
    and suggests that explaining lake levels with time (in a linear
    model) is not accurate.

-   **F-statistic**: The F-statistic tells us if the model is a
    significantly better fit than no slope (a horizontal line). With a
    p-value of 3.545e-08, the linear relationship between time and lake
    levels is highly significant.

### Residual Plots

```{r}
par(mfrow = c(2, 2))
plot(model_lake_huron)
par(mfrow = c(1, 1))

```

From the residuals vs. fitted plot, it is clear that the linear model is
not a good fit for this dataset. There should ideally be no pattern in
the residuals but we clearly see that there is a systematic pattern to
them. The red smoothed line should be approximately flat around 0 and in
the plot, it is u-curved which likely means there is a non-linear
pattern in the data. The residuals also don't have constant variance,
some are more widely spread while others are not. This suggests
heteroscedasticity and that the model is not reliable.

## Autoregression

As discussed previously, the data only has two variables. To potentially
find a model that could predict future lake levels more accurately than
a simple linear regression, we look into autocorrelation of lake levels.

Autocorrelation is the correlation of lake levels with its past lake
levels. Specifically, it measures how strongly a variable correlate with
itself at different lags (time intervals).

To examine autocorrelation, we plot the autocorrelation function.

```{r}
par(mfrow = c(1,2))
acf(LakeHuron)
curve(acf(LakeHuron)$acf[2]^x, add=TRUE, col=3, lwd=2)
par(mfrow = c(1,1))
```

The acf curve is exponential which means that autocorrelation is high in
the beginning (approximately 0.9) and steadily decreases from year to
year. Basically, the lake levels correlate more with the levels of most
recent past years. As time goes by, the autocorrelation drops which
means that past lake levels (approximately beyond 10 lags) have less
influence on current levels.

```{r}
par(mfrow = c(1, 2))
acf(LakeHuron)
curve(acf(LakeHuron)$acf[2]^x, add=TRUE, col=3, lwd=2)
par(mfrow = c(1, 1))
```

An autoregression model where we just consider the previous value (known
as AR(1)) has the formula:

$$
X_t = \phi_1 X_{t-1} + \epsilon_t
$$

-   $X_t$: lake level measured at time $t$.
-   $\phi_1$: The coefficient for the influence of the previous lake
    level value ($X_{t-1}$) on the current one ($X_t$).
-   $\epsilon_t$: Random error ($\epsilon_t \sim N(0, \sigma^2)$).

```{r}

# AR(1) linear model

ar_lake_huron <-  lm(LakeHuron[-1] ~ LakeHuron[-length(LakeHuron)])

plot(LakeHuron[-length(LakeHuron)],LakeHuron[-1])
abline(ar_lake_huron, col=3, lwd=2)

summary(ar_lake_huron)
```

There is a clear linear relationship and positive correlation between
the current lake level $X_{t}$ and last year lake level $X_{t-1}$. The
R-squared value at 0.7037 indicates that 70.37% of variation in lake
level is explained by the past year lake level. The AR(1) coefficient
$\phi_1$ of 0.83641 indicates that a one-foot increase in the previous year's lake level is associated with an approximately 0.836-foot increase in the current year's level.

Overall, the AR(1) model seems to be better suited than a simple linear
regression with time/lake level.

#### Residual diagnostic plots

```{r}
par(mfrow = c(1, 2))
# Residual diagnostic plots
plot(ar_lake_huron, 1:2)
acf(residuals(ar_lake_huron))
par(mfrow = c(1, 1))
```

The Residuals vs. Fitted plot shows no trend for the residuals and
almost a perfect fitted horizontal line at 0. This indicates that there
is no pattern for the residuals and that the AR model is a good fit.
Compared to the residuals plot of the simple linear regression, it is
clear that the AR model is the correct choice for the time series
dataset LakeHuron.

# 5: Pima Indians diabetes dataset

Women with Pima Indian heritage were tested for diabetes according to
WHO criteria. The dataset has 532 entries.

## EDA

In this task, we're fitting a logistic regression model to explain
diabetes (variable "type") using independent variables age, npreg, bmi,
glu, bp, ped and skin.

Before that, we must check if the conditions for a logistic regression
model are satisfied.

```{r}
data("Pima.tr", package = "MASS")
# remove 'type' before applying pairs() becaus it is not numerical
Pima_numeric <- Pima.tr[, sapply(Pima.tr, is.numeric)]
pairs(Pima_numeric, lower.panel = panel.smooth, diag.panel = panel.hist, upper.panel = panel.cor)

```

Key conditions:

1.  Binary dependent variable: Logistic regression is a suitable model
    because the dependent variable i.e. diabetes has only two possible
    outcomes "yes" and "no".

2.  Multicolinearity: the independent variables must not be correlated
    with each other. We check this using a correlation matrix and
    residual plots.

Looking at the correlation matrix above, we can clearly identify a
strong positive (r \> 0.5) correlaiton between: "bmi/skin" and
"npreg/age".

The strong positive correlation between "bmi" and "skin" is not
surprising as "skin" is a measure of triceps skin fold thickness in
milimiters. The correlation is also visible in the steep correlation
line on the scatterplot. We decided to remove "skin" from the rest of
the analysis as it appears to have more outliers than BMI and we
percieved BMI as a more accurate measurement of weight and height
compared to "skin" which may play a role in explaining diabetes.

Regarding the positive correlation between variables "npreg" (number of
pregnancies) and "age", we decided to remove "age" from the rest of the
analysis as it showed a mild positive correlation with both "bp" (r =
0.39) and "glu" (r = 0.34) while "npreg" didn't have mild correlations
with any other variables.

3.  No extreme outliers as logistic regression is sensitive to extreme
    values. This can be checked visually on histograms and boxplots or
    by looking at Cook's distance.

At this stage, there does not appear to be any extreme/concerning
behavior of any remaining variables in the histograms above. We will
confirm this by looking at Cook's distance in a later stage.

4.  The binary choices (yes/no) in the "type" variable should be
    approximately balanced.

```{r}
table(Pima.tr$type)
```

Looking at the balance between yes/no, we see that 34% of women had
diabetes while 66% did not. This ratio is still acceptable to continue
with fitting a logistic regression model but we can keep this
information in mind in case the model shows strong
favoring of predicting "no". In this case, weighed logistic regression
is a potential fix (assigning more weight to "yes" because it's the
minority).

## Logistic Regression Model

```{r}
model_pima <- glm(type ~ npreg + bmi + glu + bp + ped, data = Pima.tr, family = binomial)

summary(model_pima)
```

Looking at the p-values from the model, we can deduce that "bp" is not a
relevant predictor for diabetes (p = 0.79416). Hence, "bp" is removed
from the rest of the analysis. Hence, the logistic regression equation
(with variables rounded to 3 decimals) is as follows:

$$
logit(type) = \log \left( \frac{P(Y = 1)}{P(Y = 0)} \right) = -9.511 + 0.170 \cdot \text{npreg} + 0.079 \cdot \text{bmi} + 0.034 \cdot \text{glu} + 1.761 \cdot \text{ped}
$$

```{r}
model_pima_no_bp <- glm(type ~ npreg + bmi + glu + ped, data = Pima.tr, family = binomial)

summary(model_pima_no_bp)
```

Plasma glucose concentration ("glu") is a highly statistically
significant predictor in the model with a p value of **9.56e-08** . The
estimate for "glu" is **0.034749** meaning that if plasma glucose
increases by 1-unit, the logarithmic odds of having diabetes increase by
**0.034749**. If we calculate this in terms of odds ratio we get that
$exp(0.034749) \approx 1.035$, so 1-unit increase in "glu" increases
odds of having diabetes by approximately 3.5% (if all other variables
are constant).

An important predictor with a high estimate value and statistical
significance (**p = 0.00696**) is "ped" or the pedigree function
(basically the genetic risk factor based on family history of a
diabetes). The estimate for "ped" is **1.764673** meaning that log odds
for diabetes increase by this value for a 1-unit increase in "ped". The
odds ratio is $exp(1.764673) \approx 5.840$ which shows us that the odds
of having diabetes increase by around 5.8 times for a 1-unit increase in
"ped". We can conclude that family history of diabetes is a strong
predictor for having diabetes.

```{r}
format_glm_output <- function(model) {
  coef_df <- as.data.frame(summary(model)$coefficients)
  exp_values <- exp(coef_df$Estimate) 

  result <- data.frame(
    Predictor = rownames(coef_df),
    Estimate = round(coef_df$Estimate, 4),
    Exp_Estimate = round(exp_values, 4) 
  )
  print(result, row.names = FALSE)
}

format_glm_output(model_pima_no_bp)

```

The number of pregnancies ("npreg") is also a statistically sigificant
predictor (p = 0.00133) and odds ratio of $exp(0.173367) \approx 1.1893$
which means that one pregnancy increases the odds of having diabetes by
approximately 1.19 times or by about 19%.

## Cook's Distance

As the threshold we set the typical 4/n and looked at the remaining
variables to make sure there are no extreme outliers or mistakes in the
data which could be skewing the model.

```{r}
cooks_values <- cooks.distance(model_pima_no_bp)

# Identify influential points (rule of thumb: Cook's D > 4/n)
threshold <- 4 / nrow(Pima.tr)
influential_points <- which(cooks_values > threshold)
Pima.tr[influential_points, ]
```

Cook's distance identified 16 possible outliers but no value was
unusually high (as to be impossible or a clear mistake). Some women had
12 or 13 children and some had high glucose levels but all within the
expected range. Since we didn't identify any unexpected values, we
decided to proceed without removing anything.

## Receiver Operating Characteristic (ROC)

In this final part of the analysis, we use ROC to evaluate the
performance of the logistic regression model.

```{r}
#install.packages("pROC")
library(pROC)

# Predict probabilities
Pima.tr$predicted_prob <- predict(model_pima_no_bp, type = "response")

# Compute ROC curve
roc_curve <- roc(Pima.tr$type, Pima.tr$predicted_prob)

# Plot ROC curve
plot(roc_curve, col = "blue", main = "ROC Curve for Diabetes Prediction")

# Compute AUC
auc_value <- auc(roc_curve)
auc_value
```

The blue ROC-curve shows the data compared to the diagonal gray line
which is the baseline case (if the data was random). The closer the
ROC-curve to the corner, the more accurate is the model for predicting
diabetes. Vice versa, if the ROC curve is closer to the diagonal
baseline, the less accurate the model for making predictions.

To get a numerical representation of the curve, we can calculate the AUC
or area under curve which tells us how accurate the model is at
correctly recognizing a true positives and true negatives. The model
would be perfect if it had an AUC score of 1.

The **AUC score of 0.8415** indicates good discrimination between diabetic and non-diabetic cases.

### Confusion matrix

```{r}
# predicted probabilities to class labels (Threshold = 0.5)
predictions <- ifelse(Pima.tr$predicted_prob > 0.5, "Yes", "No")
predictions <- factor(predictions, levels = levels(Pima.tr$type))

# confusion matrix
conf_matrix <- table(Predicted = predictions, Actual = Pima.tr$type)
print(conf_matrix)

TP <- conf_matrix["Yes", "Yes"]
FP <- conf_matrix["Yes", "No"]
TN <- conf_matrix["No", "No"]
FN <- conf_matrix["No", "Yes"]

cat("\nConfusion Matrix \n")
cat("TP:", TP, "\n")
cat("FP:", FP, "\n")
cat("TN:", TN, "\n")
cat("FN:", FN, "\n")
```

Looking at the confusion matrix, we identify 39 true positives and 120
true negatives compared with 29 false negatives (women who had diabetes
but weren't correctly predicted) and 12 false positives.

There is a bit of concern with the false negatives as the rate of false
negatives $FNR = FN/ (FN+TP) = 0.4264706$ means that approximately 42.7%
of women who had diabetes would be incorrectly predicted as non-diabetic
by the model (and around 57.3% diabetic women would be correctly
identified). What are some possible reasons for this?

If we look back at the examination of requirements for the logistic
regression model, we found that the data contained more participants
without diabetes (66% "No"). This meant that from the beginning, the
model was slightly biased towards "No". From the confusion matrix and
false negative rate, we see that the model is moderately good at
recognizing diabetic women (**TPR = 57.3%**) and very good at
recognizing non-diabetics with a false positive rate of
$FPR = FP/(FP+TN) = 0.0909$ or just around 9.1%.

In conclusion, while the model is good at correctly predicting
non-diabetics, it has a moderate accuracy for identifying diabetics. A
possible improvement would be using weighed logistic regression to
balance out the imbalance between the number of "yes" and "no" responses
in the "type" variable
